Learning Discrete Bayesian Networks from Continuous Data

### 2016-07-28 Learning_Discrete_Bayesian_Networks_from_Continuous_Data.tex
some edits, particularly to new Discussion section

This paper describes a new Bayesian discretization method for continuous variables in Bayesian networks, extending prior work on single-variable discretization methods from \citet{Boulle_2006} and \citet{Lustgarten_2011}.
The proposed method optimizes the discretization policy relative to the network and takes parents, children, and spouse variables into account.
The optimal single-variable discretization method is derived in Section~\ref{sec:single_var}, using a prior which reduces the discretization runtime to $\bigo\paren{r \cdot n^2}$ without sacrificing optimality.
Section~\ref{sec:multi_var} covers Bayesian networks with multiple continuous variables and Section~\ref{sec:structure_learning} covers discretization while simultaneously learning network structure.
The paper concludes with a comparison against the existing minimum-description length \citep{Friedman_1996} method on real-world datasets in Section~\ref{sec:experiments}.

@@

The discretizations are compared using the mean cross validated log-likelihood of the data $D$ given the Bayesian network $B$ and discretization policies $\discset$.
Note that all log-likelihoods shown have been normalized by the number of data samples.
The log-likelihood has two components,

@@

Six additional datasets \citep{Lichman_2013} are tested to compare the Bayesian and MDL discretization methods.
The results are summarized in Table~\ref{table:more_data_set}.
Each row corresponds to a dataset and provides the number of samples $n$, the number of variables $m$, the fixed network structure, and the log-likelihood for each method.
Fixed network structures are obtained by the same principle as in previous three datasets and are represented in the table by listing the parents for each variable.
For example, in the first row, the representation $\curly{\emptyset,A,BE,AB,B,AH,BDG,B}$ indicates that $\Pa_A = \emptyset$, $\Pa_B = A$, $\Pa_c = \curly{B,E}$, etc.
The Bayesian discretization method has higher mean cross-validated log likelihood in all cases.

\begin{table}[ht]
	\centering
	\caption{
		Discretization results for several datasets.
    The number of samples, number of variables, network structure, and mean cross-validated log-likelihood under each discretization method are given.
	}
	\input{Table3_2.tex}
	\label{table:more_data_set}
\end{table}

@@

To support the argument, a synthetic dataset with $800$ samples and under the network structure in Figure~\ref{fig:exp_discuss} is tested. Data samples are generated by the distribution described in Table~\ref{table:distri_exp}: both $P_1$ and $P_2$ take two discrete values with equal probability. $\cX$ samples numeric numbers given the value of $P_1$ and $P_2$. Note that the symbol $\sim U[a,b]$ in Table~\ref{table:distri_exp} represents an uniform distribution from value $a$ to $b$. The optimal discretization policy on $\cX$ should have six edges $\curly{2.0,3.0,4.0,4.5,5.0,6.0}$, since $\cX$ takes numeric values from $1.0$ to $7.5$ and there are seven intervals corresponding to different joint distribution of $P_1$ and $P_2$.

\begin{table}[ht]
	\centering
	\caption{The distribution of variables in Figure~\ref{fig:exp_discuss} for the proposed experiment. $P_1$ and $P_2$ are discrete variable. $\cX$ is continuous valued.}
	\label{table:distri_exp}
	\begin{tabular}{cl}
		\toprule
		Variable               & Distribution                                   \\ \midrule
		\multirow{2}{*}{$P_1$}&$P(p^{(1)}_1) = 0.5$     \\
		                                       & $P(p^{(2)}_1) = 0.5$     \\ \midrule
		\multirow{2}{*}{$P_2$}&$P(p^{(1)}_2) = 0.5$     \\
		                                       &$P(p^{(2)}_2) = 0.5$     \\ \midrule
		\multirow{4}{*}{$\cX$} & $p(\cx \mid p^{(1)}_1,p^{(1)}_2) \sim U[2,6]$\\
		                                     & $p(\cx \mid p^{(1)}_1,p^{(2)}_2) \sim U[1,3]$ \\
		                                     & $p(\cx \mid p^{(2)}_1,p^{(1)}_2) \sim U[4,5]$ \\
		                                     & $p(\cx \mid p^{(2)}_1,p^{(2)}_2) \sim U[4.5,7.5]$\\
		\bottomrule
	\end{tabular}
\end{table}

Figure~\ref{fig:synthetic} shows how the likelihood of tested data and number of discretization edges on $\cX$ vary with respect to the size of trained data. Trained data are drawn randomly from the synthetic dataset. The remaining ones are tested data. If the size of the trained data is small, it is hard to recognize the underlying pattern of data samples and discretize the continuous variable $\cX$ accordingly. In this case, both likelihood of the trained data and number of discretization edges is small. However, since the Bayesian method is more sensitive to the data distribution, it has better performance of the MDL method in terms of likelihood and lays down more discretization edges. As the size of the trained data grows, both methods improve their accuracy and approach to optimal number of discretization edges. The Bayesian method finds the six discretization edges faster than the MDL method. Finally, as the size of the trained data is large enough to capture the data distribution, both methods can locate the six discretization edges and lead to same accuracy to the tested data. The curves in Figure~\ref{fig:synthetic} demonstrate that the Bayesian method is indeed more sensitive to the dataset than the MDL method. Note that all the results in Figure~\ref{fig:synthetic} are averaged over one hundred times.

@@

	\caption{The average log-likelihood of the trained data and the average number of discretization edges for $\cX$ as the number of trained data is varied. Trained data are drawn randomly from the synthetic dataset. The remaining ones are tested data. Results are averaged over one hundred times.}
	\label{fig:synthetic}
\end{figure}

>>>>

This paper describes a new Bayesian discretization method for continuous variables in Bayesian networks, extending prior work on single-variable discretization methods from \citet{Boulle_2006} and \citet{Lustgarten_2011}.
The proposed method optimizes the discretization policy relative to the network and takes parents, children, and spouse variables into account.
The optimal single-variable discretization method is derived in Section~\ref{sec:single_var}, using a prior which reduces the discretization runtime to $\bigo\paren{r \cdot n^2}$ without sacrificing optimality.
Section~\ref{sec:multi_var} covers Bayesian networks with multiple continuous variables and Section~\ref{sec:structure_learning} covers discretization while simultaneously learning network structure.
The paper concludes with a comparison against the existing minimum-description length \citep{Friedman_1996} method on real-world and synthetic datasets in Section~\ref{sec:experiments}.

@@

The discretizations are compared using the mean cross validated log-likelihood of the data $D$ given the Bayesian network $B$ and discretization policies $\discset$.
\tim{Did we also normalize each variable to be zero mean and unit covariance?}
Note that all log-likelihoods shown have been normalized by the number of data samples.
The log-likelihood has two components,

@@

Six additional datasets \citep{Lichman_2013} are tested to compare the Bayesian and MDL discretization methods.
The results are summarized in Table~\ref{table:more_data_set}.
Each row corresponds to a dataset and provides the number of samples $n$, the number of variables $N$, the fixed network structure, and the log-likelihood for each method.
Fixed network structures are obtained by the same principle as in previous three datasets and are represented in the table by listing the parents for each variable.
For example, in the first row, the representation $\curly{\emptyset,A,BE,AB,B,AH,BDG,B}$ indicates that $\Pa_A = \emptyset$, $\Pa_B = A$, $\Pa_c = \curly{B,E}$, etc.
The Bayesian discretization method has higher mean cross-validated log likelihood in all cases.

\begin{table}[ht]
	\centering
	\caption{
		Discretization results for several datasets.
    The number of samples, number of variables, network structure, and mean cross-validated log-likelihood under each discretization method are given.
	}
  \resizebox{\textwidth}{!}{
	 \input{Table3_2.tex}
  }
	\label{table:more_data_set}
\end{table}

@@

To support this argument, we test a synthetic dataset with $800$ samples using the network structure in Figure~\ref{fig:exp_discuss}.
Data samples are generated by the distribution described in Table~\ref{table:distri_exp}: both $P_1$ and $P_2$ are fair coins.
The continuous variable $\cX$ is conditionally dependent on $P_1$ and $P_2$.
Note that the symbol $\sim U[a,b]$ in Table~\ref{table:distri_exp} represents a uniform distribution from $a$ to $b$.
The optimal discretization policy on $\cX$ has six edges, $\curly{2.0,3.0,4.0,4.5,5.0,6.0}$, since $\cX$ takes numeric values from $1.0$ to $7.5$ and there are seven intervals corresponding to different joint distributions of $P_1$ and $P_2$.

\begin{table}[ht]
	\centering
	\caption{Probability distributions for the variables in Figure~\ref{fig:exp_discuss}.
           $P_1$ and $P_2$ are discrete whereas $\cX$ is continuous.}
	\label{table:distri_exp}
	\begin{tabular}{cl}
		\toprule
		Variable               & Distribution                                   \\ \midrule
		\multirow{2}{*}{$P_1$}&$P(p^{(1)}_1) = 0.5$     \\
		                                       & $P(p^{(2)}_1) = 0.5$     \\ \midrule
		\multirow{2}{*}{$P_2$}&$P(p^{(1)}_2) = 0.5$     \\
		                                       &$P(p^{(2)}_2) = 0.5$     \\ \midrule
		\multirow{4}{*}{$\cX$} & $p(\cx \mid p^{(1)}_1,p^{(1)}_2) \sim U[2,6]$\\
		                                     & $p(\cx \mid p^{(1)}_1,p^{(2)}_2) \sim U[1,3]$ \\
		                                     & $p(\cx \mid p^{(2)}_1,p^{(1)}_2) \sim U[4,5]$ \\
		                                     & $p(\cx \mid p^{(2)}_1,p^{(2)}_2) \sim U[4.5,7.5]$\\
		\bottomrule
	\end{tabular}
\end{table}

Figure~\ref{fig:synthetic} shows how the test likelihood and the number of discretization edges for $\cX$ vary with respect to the number of training samples.
Trained data are drawn randomly from the synthetic dataset.
\todo{The remaining ones are tested data.}
When the training set is small, it is difficult to recognize the underlying pattern and discretize the continuous variable $\cX$ accordingly.
In this case, both the likelihood of the training data and number of discretization edges is small.
However, the Bayesian method is more sensitive to variation in the data and produces a distribution both with higher likelihood and more discretization edges.
As the size of the training set grows, both methods converge towards the optimal discretization policy.
The Bayesian method finds the six discretization edges faster than the MDL method and consistently outperforms in likelihood.
The curves in Figure~\ref{fig:synthetic} demonstrate that the Bayesian method is indeed more sensitive to the dataset than the MDL method.
The results in Figure~\ref{fig:synthetic} are averaged over one hundred trials.

@@

	\caption{The average log-likelihood of the trained data and the average number of discretization edges for $\cX$ as the number of training samples is varied. Training data are drawn randomly from the synthetic distribution. \todo{The remaining ones are tested data.} \tim{$\leftarrow$ What does this mean?} Results are averaged over one hundred trials.
  \tim{Why do we plot training set likelihood and not test set likelihood?}}
	\label{fig:synthetic}
\end{figure}
